\vspace{-2.75cm}Digital Authoritarianism and United States Foreign Aid\vspace{-0.5cm}

### 2022-10-02 Paper.tex
Five-Pager

\documentclass{article}
\usepackage[utf8]{inputenc}
\title{\vspace{-2cm}Digital Authoritarianism and United States Foreign Aid\vspace{-0.5cm}}
\author{Nicholas Ray}
\date{\vspace{-0.5cm}September 2 2022\vspace{-1cm}}
\usepackage[margin=1in]{geometry}
\usepackage{mathtools,amssymb,amsthm}
\usepackage{setspace}
\doublespacing
\usepackage[backend=biber, style=authoryear, maxbibnames=99,uniquelist=false]{biblatex}
\renewbibmacro{in:}{}
\renewbibmacro*{volume+number+eid}{%
  \printfield{volume}
  \setunit*{\addnbthinspace}
  \printfield{number}
  \setunit{\addcomma\space}
  \printfield{eid}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
\AtEveryBibitem{
  \clearfield{issn}
  \clearfield{month}
  \clearfield{urlyear}
  \clearlist{language}
  \clearfield{note}
  \clearfield{month}
  \clearfield{day}
  \ifentrytype{online}{}{
    \clearfield{url}
  }
}
\addbibresource{ForeignAidBib.bib}
\begin{document}
\maketitle
\section*{Introduction}
The United States (U.S.) has been openly concerned about the internet since at least 1997, when President Bill Clinton exhorted the world to adopt a laissez-faire, commercially friendly approach to the internet (\cite{government1997}). By the 2000's, these preferences for a ``free'' internet were officially crystallized into foreign policy initiatives as so-called internet freedom programs cropped up in different facets of the U.S. government, such as the Department of State (e.g., \cite{henry2014}; \cite{government2021a}) and the U.S. Agency for Global Media (\cite{government2022}).

@@

Literature shows that foreign aid is typically a self-interested tool (e.g., \cite{hoeffler2011}) and that U.S. aid can be effective in stimulating liberal democratic values in a contemporary environment saturated with Chinese aid (\cite{blair2022}). Given these findings, it seems likely that a view of U.S. foreign aid over time would reflect the historical policy importance of internet freedom and the U.S.'s desire to oppose China's growing influence and authoritarian model of the internet. Specifically, I hypothesize that U.S. foreign aid to Africa is related to a recipient's internet freedom level and their influx of Chinese aid.

@@

@@

\section*{Literature Review} %improve literature review: both the number and quality of discussion on citations; it is imperative that I can claim that U.S. interest in internet freedom makes sense in terms 1) self-interest and 2) effectiveness in the current aid climate. Also, maybe contain a few citations about how the U.S. as a specific case creates foreign aid preferences (e.g., Milner and folks)
\subsection*{Determinants of Foreign Aid: Self-Interest or Altruism?}
Scholars have long debated if donors of foreign aid are motivated by altruism or self-interest. The received wisdom on foreign aid states that donors often give aid according to their strategic interests (e.g., \cite{mckinlay1977}), and economic trade has been shown to be a predictor of donor aid. Specifically, donors seem to distribute aid amongst recipients that import goods that donors have a comparative advantage in producing (\cite{younas2008}).

However, some studies have implied that altruism can be influential as well. France and Japan, for example, have displayed patterns of spending associated with past colonial ties and UN votes, respectively (\cite{alesina2000}). Recipient humanitarian need is also thought to be important in determining aid flows and that, despite inefficiencies in allocation, aid still results in substantial poverty reduction (\cite{collier2002}). However, more recent work on recipient need has seemed to conclude that it is largely dominated by donor self-interest, although donor behavior is still poorly understood (\cite{hoeffler2011}.

There is also seeming variation amongst donor countries depending on if they are a part of the Development Assistance Committee (DAC). More traditional donors that operate with DAC are shown to care more for recipient need while nonDAC donors allocate aid more so with regard to self-interest (\cite{dreher2011}). Generally, though, DAC aid providers care less for recipient need than previously thought while nonDAC members are less self-interested than expected by previous literature.

\subsection*{Foreign Aid: An Effective Tool?}
Regardless of the motivations behind foreign aid, there has been a significant amount of disagreement over whether foreign aid is effective in achieving its objectives. Aid, for example, has been found to be ineffective in altering levels of inequality or poverty in recipient countries, even if those countries have relatively strong institutions (\cite{chong2009a}). The lack of intentionality behind aid allocation is thought to contribute to its ineffectiveness, with donors not necessarily interested in maximizing aid effectiveness on poverty reduction or humanitarian need (\cite{alesina2000}, \cite{collier2002}).

There is, however, modest evidence that economic aid contributes to democratic consolidation in Africa (\cite{dietrich2015a}).

Interestingly, evidence for the effectiveness of aid appears to have increased with China's monumental entrance into aid provision. For example, when comparing the effects of Chinese and U.S. aid in Africa, U.S. aid is relatively successful in promoting liberal democratic values amongst Afrobarometer respondents (\cite{blair2022}). Chinese aid, by contrast, largely has null or even positive effects on democratic values as well.

Further research on aid's impact on democratic preferences in Africa after the rise of Chinese aid finds that aid from the World Bank similarly increased liberal democratic attitudes (\cite{gehring2022}). World Bank aid is even associated with improved stability, whereas Chinese aid is shown to have relatively smaller effects on stability and a positive relationship to autocratic acceptance.

There is a substantial amount of nuance to aid effectiveness, though. Aid from China is associated with positive economic growth in recipient countries while U.S. aid is relatively more effective in countries not currently receiving Chinese aid (\cite{dreher2021}).

Other work details how bilateral aid is more likely to be conditioned on levels of corruption and thus more effective than multilateral aid (\cite{christensen2011}).

\section*{Methodology}
Following the logic and stated hypothesis in the introduction, the dependent variable is U.S. foreign aid to countries in Africa while the main explanatory variables are recipient country internet freedom levels and Chinese foreign aid to Africa. The time-span of the data range from 2000-2017, with the data described as follows.

U.S. foreign aid is measured using disbursement data from the U.S. government (\cite{government2022e}), which reports foreign aid on a fiscal-year basis. The measurement for Chinese foreign aid stems from the College of William \& Mary (\cite{custer2021}) which quantifies official financial and in-kind commitments made by China. Internet freedom data is provided by the Digital Society Project (DSP, \cite{mechkova2022}), being collected through expert surveys and estimated with methodology designed by the Varieties of Democracy project (V-Dem, \cite{coppedge2022}). The DSP dataset is potentially superior for measuring the concept of internet freedom compared to the alternative, Freedom of the Net (FOTN, \cite{house2022}). FOTN features frequent missing data for the continent of Africa and does not take into account possible variations in expert opinion and knowledge like the DSP does.

According to the most recent advice for determining good controls (\cite{cinelli2022}), I include GDP per capita, an indicator for internet development and political freedom as controls. GDP per capita is assessed in terms of 2021 U.S. dollars as is supplied by the World Bank (\cite{bank2022}). The indicator for internet development, internet usage by country, is calculated as a percentage and is reported by the International Telecommunication Union (ITU, \cite{itu2022}). Lastly, the chosen measure for political freedom comes from V-Dem (\cite{coppedge2022a}).

OLS with country fixed effects

IV?

Case study comparing U.S. aid changes to countries that had similar internet freedom levels but experienced different influxes of Chinese aid.

\section*{Limitations}
Definition of internet freedom (difference between DSP, U.S.)

Endogeneity.

\section*{Conclusion}
Restate important parts of the introduction.

Speculate on theoretical reasons why U.S. aid might or might not be related to internet freedom and Chinese foreign aid. If positively related, say that internet freedom is good for capitalist business. If null, perhaps data doesn't capture recent trends as focus has increased. Also, perhaps government decision making slow to react (Congress) or unwilling for domestic reasons (Milner stuff, for example).

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\title{\vspace{-2.75cm}Digital Authoritarianism and United States Foreign Aid\vspace{-0.5cm}}
\author{Nicholas Ray}
\date{\vspace{-0.30cm}October 2 2022\vspace{-1cm}}
\usepackage[margin=1in]{geometry}
\usepackage{mathtools,amssymb,amsthm}
\usepackage{setspace}
\doublespacing
\usepackage[backend=biber, style=authoryear, maxbibnames=99,uniquelist=false]{biblatex}
\renewbibmacro{in:}{}
\renewbibmacro*{volume+number+eid}{%
  \printfield{volume}
  \setunit*{\addnbthinspace}
  \printfield{number}
  \setunit{\addcomma\space}
  \printfield{eid}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
\AtEveryBibitem{
  \clearfield{issn}
  \clearfield{month}
  \clearfield{urlyear}
  \clearlist{language}
  \clearfield{note}
  \clearfield{month}
  \clearfield{day}
  \ifentrytype{online}{}{
    \clearfield{url}
  }
}
\addbibresource{ForeignAidBib.bib}
\begin{document}
\maketitle
\section*{Introduction}
The United States (U.S.) has been openly concerned about the internet since at least 1997, when President Bill Clinton exhorted the world to adopt a laissez-faire, commercially friendly approach to the internet (\cite{government1997}). By the 2000's, these preferences for a ``free'' internet were officially crystallized into foreign policy initiatives as so-called ``internet freedom'' programs cropped up in different facets of the U.S. government, such as the Department of State (e.g., \cite{henry2014}; \cite{government2021a}) and the U.S. Agency for Global Media (\cite{government2022}).

@@

Literature shows that foreign aid is typically a self-interested tool (e.g., \cite{hoeffler2011}) and that U.S. aid can be effective in stimulating liberal democratic values in a contemporary environment saturated with Chinese aid (\cite{blair2022}). Given these findings, it seems likely that a view of U.S. foreign aid over time would reflect the historical policy importance of internet freedom and the U.S.'s desire to oppose both China's growing influence and their authoritarian model of the internet. Specifically, I hypothesize that U.S. foreign aid to Africa is related to a recipient's internet freedom level and their influx of Chinese aid.

@@

Understanding the U.S.'s actualization of internet preferences has important implications not only for the foreign aid literature, but also for international relations and comparative scholars more broadly. Firstly, any result that indicated the importance of internet considerations in inter-state behavior would be the first of its kind to my knowledge, and would signify a potentially novel dimension along which states interact. Secondly, this paper would help inform the dynamics of the U.S. and China's ongoing contest for economic and political influence by highlighting another issue area by which they compete. Lastly, this work would raise questions about the generalizability of state concern over the internet, which could encompass studies on conflict and cyberspace, research on democratization and information asymmetries, and political economic work on globalization.

@@

\section*{Literature Review}
\subsection*{Altruistic versus Self-Interested Aid}
Scholars have long debated if donors of foreign aid are motivated by altruism or self-interest. The received wisdom on foreign aid states that donors often give aid according to their strategic concerns (e.g., \cite{mckinlay1977}; \cite{alesina2000}), and economic trade has more recently been shown to be a predictor of donor aid. Specifically, donors seem to distribute aid amongst recipients that import goods that donors have a comparative advantage in producing (\cite{younas2008}).

However, some studies have implied altruism is also an important factor in determining aid flows, especially for private aid and non-governmental organizations (\cite{buthe2012}). Other scholars have analyzed the potential for aid to reduce poverty, concluding that poverty reduction is occurring but in an inefficient manner (\cite{collier2002}). The consensus, though, still seems to be that recipient humanitarian need is largely dominated by donor self-interest (e.g., \cite{hoeffler2011}).

The balance between strategic interests and altruism has additionally been conceived of along the lines of donor maturity, with more established donors potentially distributing aid in more or less altruistic ways than ``new'' donors (\cite{dreher2011}). Generally, though, it is found that traditional aid providers care less for recipient need than previously thought while relatively young donors are less self-interested than expected by previous literature.

\subsection*{Foreign Aid: An Effective Tool?}
Regardless of the exact source of motivation behind foreign aid, there has been a significant amount of disagreement over whether foreign aid is effective in achieving any objectives. In some cases, for example, it appears that the lack of discernible and consistent motivations is the very reason why aid can be found to be ineffective (\cite{collier2002}). Aid has also been found to be ineffective in altering levels of inequality or poverty in recipient countries, even if those countries have relatively strong institutions (\cite{chong2009a}).

There is, however, modest evidence that economic aid effectively contributes to democratic consolidation in Africa (\cite{dietrich2015a}). Effectiveness can vary by certain characteristics of aid as well, with bilateral aid being more likely to be conditioned on a recipient country's level of corruption. Multilateral aid, on the other hand, is generally more impartial and unconditional, resulting in diminished effectiveness when compared to bilateral aid (\cite{christensen2011}).

Interestingly, evidence for the effectiveness of aid appears to have strengthened with China's monumental entrance into the aid arena. For example, when comparing the effects of Chinese and U.S. aid in Africa, U.S. aid is found to be relatively successful in promoting liberal democratic values amongst Afrobarometer respondents (\cite{blair2022}). Chinese aid, by contrast, largely has null or even similarly positive effects on democratic values. In a separate study relating World Bank and Chinese aid, foreign aid from the World Bank is associated with the same liberal democratic values and even improved country stability. A feeling of acceptance towards autocratic leadership is found to be connected to Chinese aid (\cite{gehring2022}).

Yet there is a substantial amount of nuance to aid effectiveness. Aid from China is definitively linked with positive economic growth in recipient countries, for example, while U.S. aid is most effective in countries not currently receiving large amounts of Chinese aid (\cite{dreher2021}).

Thus, it can be gathered from previous literature that donor behavior, including that of the U.S., should be expected to be a product of strategic considerations. Furthermore, while not always effective in securing these considerations, it is plausible that the current atmosphere of aid-giving in Africa represents a scenario in which aid tied to liberal democratic ideals can be effectual.

\section*{Empirics}
\subsection*{Data and Variables}
In approaching an analysis of the relationship between foreign aid and internet freedom, it perhaps makes most sense to firstly define the main explanatory variable of internet freedom. The Digital Society Project (DSP) aims to identify ``how people use social media as a political tool and to explore how political institutions and social media use interact,'' providing data from expert surveys that cover a range of topics such as censorship, disinformation, and monitoring (\cite{mechkova2022}). The DSP employs methodology designed by the Varieties of Democracy project (V-Dem, \cite{coppedge2022}) to try and estimate differences in expert opinion and knowledge.

Data from the DSP is potentially superior for measuring the concept of internet freedom compared to the primary alternative, Freedom on the Net (FOTN, \cite{house2022}). FOTN features frequent missing data for the continent of Africa and does not take into account possible variations in expert opinion and knowledge like the DSP does.

The other main explanatory variable is Chinese foreign aid, stemming from the College of William \& Mary's AidData research lab (\cite{custer2021}). This is the one of the only datasets on Chinese aid and is certainly the largest and most specific. Aid is quantified as official financial and in-kind commitments and is measured in constant 2017 USD.

U.S. foreign aid is the dependent variable, measured as fiscal-year disbursements (\cite{government2022e}). It is measured in constant 2022 USD.

The starting year for the holistic dataset is 2000, defined by the lower bound of data availability for internet freedom, and 2017 is the ending year per data availability for Chinese foreign aid. There is some evidence of left-truncation or missing data in the information on internet freedom, but otherwise all African countries are accounted for.

\subsection*{Methodology}
To estimate the association between U.S. foreign aid and the main explanatory variables, I plan to use ordinary least squares (OLS) with country fixed-effects. Fixed-effects at the country level help account for any underlying similarity between countries that could induce spuriousness.

In addition to including fixed-effects, there are multiple controls that seem appropriate to incorporate. According to the most recent advice for determining good controls (\cite{cinelli2022}), I include gross domestic product (GDP) per capita, an indicator for internet development, and political freedom as controls. Each of these phenomena are possible confounders, being theoretically related to at least one of the types of aid and internet freedom.

GDP per capita is assessed in constant 2021 USD and as is supplied by the World Bank (\cite{bank2022}). The indicator for internet development, internet usage by country, is calculated as a percentage and is provided by the International Telecommunication Union (\cite{itu2022}). Lastly, the chosen measure for political freedom comes from V-Dem (\cite{coppedge2022a}).

\section*{Limitations}
There are at least a couple important limitations to be clear about. Firstly, there is undoubtedly some measurement error in the quantification of internet freedom brought about due to possible differences in the idea of internet freedom employed by the U.S. and DSP. In other words, even if the U.S.'s conception of internet freedom was constant over time, the DSP definition is likely to be somewhat distinct.

Secondly, there is potential simultaneity between U.S. foreign aid and internet freedom. More advanced econometric modeling techniques, such as difference-in-differences (DiD) or instrumental variables regression (IV) may be imperative here.

\section*{Conclusion}
It is possible that there is no meaningful association between U.S. foreign aid and internet freedom. The lack of association would in itself present a sort of puzzle, though, since the U.S. clearly has a foreign policy interest in maintaining internet freedom and a seemingly effective tool of foreign aid to pursue that interest. It may take time, however, for the ever-increasing advances in U.S. financial commitment to internet freedom to be reflected in spending data.

If there is indeed a relationship, I speculate that the U.S. is interested in keeping the internet free because it is good for business. There is evidence that China is sharing it's technology, like censorship and monitoring tools, that could allow more authoritarian-leaning developing countries to prevent the use of U.S. technology and to strictly allow for Chinese substitutes instead. These developing countries would then harness the economic benefits of the internet, like job growth, while constraining potentially negative political externalities such as the free assembly of anti-incumbent online groups. China would stand to gain domestic economic growth as it's technology firms expand beyond it's borders, and everybody wins- except the U.S.
